---
title: "AdvancedRNA-seq"
author: "Simon Tomlinson"
date: "16/02/2026"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of RNA-seq data using R 

•	You should have created a new folder for this tutorial, copied and unpacked  the tutorial files 
•	Now run this script


```{r init, echo=FALSE}
#Load the required libraries & load the files for the tutorial
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(QuasR)
library(pheatmap)
load("FGT_T7_RNA-seq.Rdata")
#We now have files for the tutorial and also data loaded into the workspace- use ls() and dir() to view this data
ls()
dir()


```
Once we have loaded the data we are ready to run some further analysis.

```{r input_processing, echo=FALSE}
#Some further processing to access count data and column names.
table_rnaseq<-(mytable_feaures)$counts
colnames_rnaseq <-colnames(table_rnaseq)
#tidy up the sample names eg"X.home.guest101.new_rnaseq.aligned.SRR5054353.sorted.bam"
#to ShortName  using the “ targets file called adf here
adf <-read.table("mychromosome_samples4.txt",sep='\t', row.names=1,fill=T,header=T)
#This orders the elements in the table
idx<-match(colnames_rnaseq,rownames(adf))
colnames(table_rnaseq) <-adf$ShortName[idx]
adf <-adf[idx,]
rownames(adf) <-colnames(table_rnaseq)
```
## Exploring the Sample Annotation
We can use R to summarise the annotation table from this study. 
```{r explore_annot, echo=FALSE}

#Load the annotation table- here it is a data frame
adf <-read.table("mychromosome_samples4.txt",sep='\t', row.names=1,fill=T,header=T)
#The annotation table contains data in a number of named columns
colnames(adf)
#We can generate a histogram to explore the library sizes in this data.
hist(as.numeric(adf$Bases.M.))
#It is also interesting to summarise the table in terms of factors called “Line” and “Sex”.  
table(adf[,c("Line","Sex")])

```

## Counting Genomic Features
We now have reads mapped to the genome, gene annotation files but the reads are not mapped and counted within each annotation region (eg within each gene).  The reads aligned to the genome are in a file format called BAM see  https://www.ncbi.nlm.nih.gov/sra/docs/submitformats/.  If we were to view this file we would see the regions to which the reads map, but in the next stage of the pipeline we use a tool called featureCounts to count the reads that map within each region we are interested in.  In the interest of time, the command is provided but commented out.  The results of running featureCounts were loaded into R when we loaded the serialised “.Rdata” object for this experiment.

```{r counting_genomic_features, echo=TRUE}
#use featureCounts to load the BAM files into the 
#mytable_feaures <-featureCounts(files=alignment_files,annot.ext="./omes/annot/Mus_musculus.GRCm38.99.gtf",isGTFAnnotationFile = TRUE, countMultiMappingReads=F, minMQS=30, nthreads=10,minOverlap=10)
#feature counts takes the genome annotation file #(Mus_musculus.GRCm38.99.gtf) and the BAM files themselves.  Note #that alignment_files is just a list of file locations.
#Check the contents of the alignment_files
alignment_files
#Note also that we have countMultiMappingReads =F this means that #reads that map to more than one genome location are ignored.
#minMQS – minimum mapping quality score,30 is a very stringent value #(we return to this later)
#nthreads =10 run ten threads at once (another reason to pre-compute #this in a large class!
#minOverlap =10 – the minimum overlap between the feature and the read- here 10bp 
#Some further processing to access count data and column names.
table_rnaseq<-(mytable_feaures)$counts
colnames_rnaseq <-colnames(table_rnaseq)
#tidy up the sample names #eg fix #"X.home.guest101.new_rnaseq.aligned.SRR5054353.sorted.bam"!
#to ShortName using the “ targets file” object called adf here
#This orders the elements in the table
idx<-match(colnames_rnaseq,rownames(adf))
colnames(table_rnaseq) <-adf$ShortName[idx]
adf <-adf[idx,]
rownames(adf) <-colnames(table_rnaseq)
#Finally check the output of the reorganisation
head(table_rnaseq)


```


## Removal of Unwanted Samples
Notice from the annotation we have two XO samples, we can convert the design to a two factor factorial design by removing these two samples.  We then have three samples in each combination of the two factors.


```{r remove_unwanted_samples, echo=TRUE}
#Adjust sample data
table(adf[,c("Line","Sex")])
#keep-remove XO samples
keep_cols=adf$Sex[idx]!="O"
#We can use this to remove the unwanted columns from the adf and #also the data eg
table(adf[,c("Line","Sex")])
#With just the required columns
table(adf[keep_cols,c("Line","Sex")])
adf <-adf[keep_cols,]

```

## Loading the Data into DESeq
DESeq2 will used for the statistical analysis and the normalisation of the data. 
```{r loading_deseq, echo=FALSE}
#Create object for DESeq
#NOTE the design for differential analysis is fed into this object 
dds_rnaseq <- DESeqDataSetFromMatrix(countData = table_rnaseq[,keep_cols],colData = adf,design = ~ Line + Sex)
#check the object dimensions
dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)
#check the object type
class(dds_rnaseq)
typeof(dds_rnaseq)

#check what this object contains
slotNames(dds_rnaseq)
colnames(colData(dds_rnaseq))
colnames(assay(dds_rnaseq))

```

Quality Filtering Genes
Apply filters to remove genes that do not pass quality filters.


```{r apply_filters, echo=FALSE}
#first filter based upon counts
keep <- rowSums(counts(dds_rnaseq)) > 1
dds_rnaseq <- dds_rnaseq[keep,]
nrow(dds_rnaseq)
#second filter we might apply-require >10 in >= 3 samples
#rowSums here counts the number of TRUE elements per row
# counts(dds_rnaseq) >= 10 returns TRUE if each row has >= 10 TRUE
#counts 
keep <- rowSums(counts(dds_rnaseq) >= 10) >= 3
dds_rnaseq <- dds_rnaseq[keep,]
#We now have a DESeq object filtered for samples and genes that pass #QC
nrow(dds_rnaseq)
```

## Normalisation of Data


```{r normalisation, echo=TRUE}
#size factor estimation for normalisation- these are used to scale by library size when data is exported.
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)
#The size factors should relate to the library size.  We can check this from the annotation
adf_bases <- adf$Bases.M.
names(adf_bases) <- rownames(adf)
adf_bases <- adf_bases/mean(adf_bases)
#Check how these two tables relate
plot(sizeFactors(dds_rnaseq), adf_bases)

```

## DESeq2 also provides methods that outputs data with various normalisations.
Here we generate multiple different normalisations and compare between then- but for the rest of the tutorial we use rlog normalisation!  Note that the normalised data is not used for differential expression analysis- just for data visualisation.


```{r plot_normalised, echo=FALSE}

#Normalised log counts
counts_rnaseq <-log2(counts(dds_rnaseq, normalized=TRUE))
#fpm is similar to cpm and discussed here https://rdrr.io/bioc/DESeq2/man/fpm.html.  Here the data is log transformed
fpm_rnaseq <-log(fpm(dds_rnaseq))#this is also normalised by the sizeFactors
#rlog ad vsd are discussed here https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8
vsd_rnaseq <- vst(dds_rnaseq, blind = T)
head(vsd_rnaseq, 3)
rld_rnaseq <- rlog(dds_rnaseq, blind = T)
head(rld_rnaseq, 3)


```
We will use rlog for the rest of this tutorial, but here we compare the different normalised data

##Plots with Normalised Data.

```{r build_plots, echo=FALSE}
boxplot(counts_rnaseq)
boxplot(fpm_rnaseq)
boxplot(assay(vsd_rnaseq))#etc
boxplot(assay(rld_rnaseq))#etc
#but suppose we look at non-normalised
na.rm=T
counts_un <-counts(dds_rnaseq, normalized=FALSE)
boxplot(log2(counts_un))
counts_un_rlog<-rlog(counts_un)
#We can make an mva.pairs plot for any of these data
#For example we can examine the unnormalized data.
#make mva.pairs-this is slow “mva.pairs.png” is provided 
#png(filename="mva.pairs.png",width=2000, height=2000)
#mva.pairs(counts_un_rlog)
#dev.off()

```

Exploratory Data Plots
PCA plots similar to those we use for microarrays.


```{r more_explore, echo=FALSE}
#we can run a PCA plot here...
library(scatterplot3d)
library(ggplot2)
# Perform PCA
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale=T)
# Plot the PCA results
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=colData(dds_rnaseq)$Colour)
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq),pos = 3,offset = 0.5,cex=0.5)
#or a quick plot in 2d
qplot(pca$x[,1],pca$x[,2],xlab="PCA1", ylab="PCA2",color=colData(dds_rnaseq)$Colour)
#We can also output a distance matrix…
sampleDists <- dist(t(assay(rld_rnaseq)))
sampleDists

```
## Generate further heatmaps with the data

```{r more_heatma-s, echo=FALSE}
library("pheatmap")
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix)
#easier PCA plot
plotPCA(rld_rnaseq, intgroup = c("Sex", "Line"))
```


## Differential Gene Expression
Here we perform a two factor analysis using DESeq2.  The design was fed into the dds_rnaseq object and was specified out as  ~Line + Sex which specifies two factors Sex and Line.   Sex and Line are colnames in the annotation able and samples are therefore annotated with these labels.  Statistics in DESeq2 works in a similar way to Limma but is designed for count data. 

```{r differential_expression, echo=FALSE}
dds_rnaseq <- DESeq(dds_rnaseq)
#This is how you specify a contrast…
result_sex <-results(dds_rnaseq,contrast=c("Sex","M","F"))
result_cell_line <-results(dds_rnaseq,contrast=c("Line","BC","CB"))
#make a nice plot of differential expression
plotMA(result_sex, main="DESeq2 dispersion plot", ylim=c(-2,2))
table(result_sex$padj < 0.01)
table(result_cell_line$padj < 0.01)
table(result_sex$padj < 0.05)
table(result_cell_line$padj < 0.05)
result_sex_selected <- subset(result_sex, padj < 0.05)
result_sex_selected <-result_sex_selected[order(abs(result_sex_selected$log2FoldChange),decreasing = TRUE), ]
head(result_sex_selected)
```

## Make heatmaps from each of the output tables- selected top 50 or all using fold change and FDR.

```{r differential_expression_heatmaps, echo=FALSE}
top50 <-rownames(result_sex_selected)[1:50]
#make a heatmap for the top 50
pheatmap(assay(rld_rnaseq)[top50,],scale="row",show_rownames=F,main="RLD values, Row Scaled, Top 50")
#all
pheatmap(assay(rld_rnaseq)[rownames(result_sex_selected),],scale="row",show_rownames=F,main="Sex Selected, Row Scaled")
#and the cell lines
result_cell_selected <- subset(result_cell_line, padj < 0.05)
result_cell_selected <-result_cell_selected[order(abs(result_cell_selected$log2FoldChange),decreasing = TRUE), ]
head(result_cell_selected)
top50c <-rownames(result_cell_selected)[1:50]
#make a heatmap for the top 50
pheatmap(assay(rld_rnaseq)[top50c,],scale="row",show_rownames=F, main="Cell Line Selected, Row Scaled, Top 50")
#all
pheatmap(assay(rld_rnaseq)[rownames(result_cell_selected),],scale="row",show_rownames=F,main="Cell Line Selected, Row Scaled")
```


## Build Gene Annotation. A serialised result "mart" has been generated in case Ensembl is unavailable

```{r build_gene_annotation, echo=TRUE}
#Download Ensembl annotation using BiomaRt and rename the samples
library(biomaRt)
#If UK ensembl is slow we can also use a USA mirror, "useast.ensembl.org"
ensembl_host <-"https://www.ensembl.org"
head(biomaRt::listMarts(host = ensembl_host), 15)
head(biomaRt::listAttributes(biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart = useMart("ENSEMBL_MART_ENSEMBL",host = ensembl_host))), 40) 
mart <- biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart    = useMart("ENSEMBL_MART_ENSEMBL",host    = ensembl_host))
resultAnnot <- biomaRt::getBM(values=rownames(dds_rnaseq),attributes = c("ensembl_gene_id","external_gene_name","chromosome_name","start_position","end_position","description","strand"),filters="ensembl_gene_id",mart=mart)

```

Merge Annotation with Input Data
```{r merge_gene_annotation, echo=FALSE}
#merge with input data
names <- resultAnnot[,1]
resultAnnot <- as.data.frame(resultAnnot)
rownames(resultAnnot) = names
idx<-match(rownames(dds_rnaseq),rownames(resultAnnot))
#make sure annotation is in same order
all(rownames(dds_rnaseq) == rownames(resultAnnot))
grr<-resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id),]
all(rownames(dds_rnaseq) == rownames(grr))
resultAnnot <-grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#make the nice names
nice_names<-paste(resultAnnot$ensembl_gene_id,resultAnnot$external_gene_name, sep = '_')
resultAnnot$nice_names <-nice_names
head(resultAnnot)
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#check names
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)#
idx2 <-match(rownames(result_sex_selected)[1:50],rownames(dds_rnaseq))
plotme <-(rld_rnaseq)[rownames(result_sex_selected)[1:50],]
rownames(plotme)<-resultAnnot$nice_names[idx2]
#make heatmap with candidate genes
png(filename="heatmap_candidates.png",width=800, height=1000)
pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40)
dev.off()

pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40)

```


## Marker gene Profiles

```{r marker_gene_profiles, echo=FALSE}
#But what about the marker profiles from known markers?
#Load some developmental markers to assess the expression profile
mkr <-read.table("markers.txt",stringsAsFactors = F,fill=T,row.names = 1,header=T,sep="\t")
rownames(mkr)
#%in%  This makes sure the marker gene names are rownames in the object
valid_names <-rownames(mkr)[rownames(mkr) %in% rownames(dds_rnaseq)]
plotmetoo <-rld_rnaseq[valid_names,]
idx3 <-match(rownames(plotmetoo),rownames(dds_rnaseq))
rownames(plotmetoo)<-resultAnnot$nice_names[idx3]

knownmarkerplot <-pheatmap(assay(plotmetoo),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40,main="Known Markers")

#make a standalone version
png("plotofknownmarkers.png", width=600,height=600)
knownmarkerplot
dev.off()

```

##Finally add column annotation

```{r column_annotation, echo=FALSE, out.width="100%", out.height="100%"}
#put in column annotation
mkr_bit <-mkr[valid_names,]
rownames(mkr_bit) <-resultAnnot$nice_names[idx3]

Sys.setlocale("LC_ALL", "C")
plotcellstate <-pheatmap(assay(plotmetoo),scale="row",fontsize_row = 10,annotation_row=as.data.frame(mkr_bit[,c("LongState","ShortState")]),cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40, main="Plot with Cell Type Information",cluster_rows = TRUE)

#if it does not fit make a png
png("plotofcelltype.png", width=600,height=600)
plotcellstate
dev.off()

#Perhaps make a plot ordered by the factor so that patterns are easier to see
mkbit_ordered <-mkr_bit[order(mkr_bit$ShortState),]
plotmarkers <-assay(plotmetoo)
plotmarkers <-plotmarkers[rownames(mkbit_ordered),]

plotcellstate_ordered <-pheatmap(plotmarkers,scale="row",fontsize_row = 10,annotation_row=as.data.frame(mkbit_ordered[,c("LongState","ShortState")]),cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40, main="Plot with Cell Type Information Factor Ordered",cluster_rows = FALSE)

#if it does not fit make a png
png("plotofcelltype_ordered.png", width=600,height=600)
plotcellstate_ordered
dev.off()




```

Additional Reading
http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

## Session Information
```{r session_info, echo=FALSE}

sessionInfo()

```


